Fitting the Undetected: Light-curve Feature Estimation in the Presence of Non-detections

### 2013-03-27 tex/ms.tex
more intro text

Telescopes don't make catalogs, they make images (Hogg \& Lang 20xx).
Nonetheless, some of the most important astronomical projects have produced enormously productive catalogs
(the \textsl{SDSS}, for examle CITATION).
There is a growing understanding that precise measurements of astronomical sources have to be made directly in the imaging data
(CITE Lang, Anderson, something in weak lensing):
It is impossible to pass the full information content of the imaging to the end user via a catalog
(which by necessity is built with point estimates of the quantities of interest).

@@

The benefit of using the non-detections is large, but the cost is also large:
Use of these ``null'' data involves non-trivial modeling,
and large numbers of latent parameters to be inferred and marginalized out.

To make this all concrete, consider the photometric data from the \textsl{ASAS} survey
depicted in \figurename~\ref{fig:example} for one Mira variable.
The photometric detections are shown and so are the epochs at which there are nulls (or non-detections).
The standard

...as an example, consider \figurename~\ref{fig:example}...

\joey{What people actually do, what problems that could cause, and what people should do}

A photometric survey (think of it as operating in a single photometric
bandpass for now) scans over the celestial position of a particular
variable star  a large number of times $t_i$.  At each of these
times, the imaging data are analyzed by software, which
treats each scan as a \emph{completely independent} survey.  That is,
when analyzing the data from time $t_i$, none of the information from
any of the other times is used in any way.

At each time $t_i$, the star is either detected by the
software or not.  When it is detected, the software returns
a (possibly bad) flux value $f_i$ and (likely bad) uncertainty
variance $s_i^2$.  When the source is \emph{not} detected,
nothing is reported---in this case, because the software is not
awesome, it doesn't see any reason to report anything at all, since on
the non-detect passes, it has no inkling that this patch of the sky is
interesting in any way.  Welcome to the desert of the real.

Beyond this, there are two additional issues.  The first is that each
night of imaging is different in an unreported and unknown way.  That
is, there is different transparency, sky brightness, and point-spread
function.  So the detection limit, or completeness level, or censoring
of the catalog is different every night.  The second issue is that
either because we are getting the data from a non-generous source, or
because the conditions change rapidly and unpredictably, we can't
analyze all the sources in a finite patch simultaneously.  For each
variable star of interest, we \emph{only} get the data on that star
itself.

>>>>

Telescopes don't make catalogs, they make images (Hogg \& Lang 20xx).
Nonetheless, some of the most important astronomical projects have produced enormously productive catalogs
(the \textsl{SDSS}, for example CITATION).
There is a growing understanding that precise measurements of astronomical sources have to be made directly in the imaging data
(CITE Lang, Anderson, something in weak lensing):
It is impossible to pass the full information content of the imaging to the end user via a catalog
(which by necessity is built with point estimates of the quantities of interest).

@@

To make this all concrete, consider the photometric data from the \textsl{ASAS} survey
depicted in \figurename~\ref{fig:badfit} for one Mira variable.
The photometric detections are shown and so are the epochs at which there are nulls (or non-detections).
The standard fitting procedure, given censored data like these, is to \emph{ignore the null values}.
When the nulls are ignored, the period fitting finds a short period.
At that period, the model predicts the star to be bright at many phases in which the survey reported nulls.
That is, although we have no idea how or why the nulls were generated,
they effectively provide strong evidence against the short period.
When we include the nulls (as described below), we get a longer period.
This longer period is obviously better because it makes the star faint whenever nulls are common.
This example clearly illustrates that the nulls are informative.
In this paper we demonstrate how to automatically extract that information.

The benefit of using the non-detections is large, but the cost is also large:
Use of these ``null'' data involves non-trivial modeling,
and large numbers of latent parameters to be inferred and marginalized out.
In particular it involves modeling the data-generating process,
which includes variable instrument and atmospheric conditions
(leading to different effective detection thresholds)
and also the software process by which catalog-entry decisons are made.
In practice our models will be very simplified, but it is impossible to proceed \emph{without} significant modeling.

\hogg{either merge in or delete the material below this line}
